---
title: "Lab 3A: Starting with the Data"
author: "Jarrod Daniels"
date: "9/7/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Presentation and Introduction of Survey Data

We are going to use the R function download.file() to download the CSV file that contains the survey data from Figshare, and we will use read.csv() to load into memory the content of the CSV file as an object of class data.frame. Inside the download.file command, the first entry is a character string with the source URL (“https://ndownloader.figshare.com/files/2292169”). This source URL downloads a CSV file from figshare. The text after the comma (“data_raw/portal_data_joined.csv”) is the destination of the file on your local machine. You’ll need to have a folder on your machine called “data_raw” where you’ll download the file. So this command downloads a file from Figshare, names it “portal_data_joined.csv” and adds it to a preexisting folder named “data_raw”.
```{r downloading data, eval=FALSE}
download.file(url = "https://ndownloader.figshare.com/files/2292169",
              destfile = "data/portal_data_joined.csv")
```


Once downloaded, we can load the data
```{r reading csv}
surveys <- read.csv("data/portal_data_joined.csv")
```

This statement doesn’t produce any output because, as you might recall, assignments don’t display anything. If we want to check that our data has been loaded, we can see the contents of the data frame by typing its name: surveys.

Wow… that was a lot of output. At least it means the data loaded properly. Let’s check the top (the first 6 lines) of this data frame using the function head():


```{r head}
head(surveys)
```

```{r view}
View(surveys)
```

## What are data frames?

Data frames are the de facto data structure for most tabular data, and what we use for statistics and plotting.

A data frame can be created by hand, but most commonly they are generated by the functions read.csv() or read.table(); in other words, when importing spreadsheets from your hard drive (or the web).

A data frame is the representation of data in the format of a table where the columns are vectors that all have the same length. Because columns are vectors, each column must contain a single type of data (e.g., characters, integers, factors). For example, here is a figure depicting a data frame comprising a numeric, a character, and a logical vector.

```{r structure}
str(surveys)
```

# Challenge Question 1
Based on the information given to us by the structure function above, we know that for surveys...
The class is a data frame
There are 34,786 rows with 13 columns
There were 48 unique species recorded during these sruveys


## Indexing and subsetting data frames
Our survey data frame has rows and columns (it has 2 dimensions), if we want to extract some specific data from it, we need to specify the “coordinates” we want from it. Row numbers come first, followed by column numbers. However, note that different ways of specifying these coordinates lead to results with different classes.
```{r indexing}
# first element in the first column of the data frame (as a vector)
surveys[1, 1]   
# first element in the 6th column (as a vector)
surveys[1, 6]   
# first column of the data frame (as a vector)
surveys[, 1]    
# first column of the data frame (as a data.frame)
surveys[1]      
# first three elements in the 7th column (as a vector)
surveys[1:3, 7] 
# the 3rd row of the data frame (as a data.frame)
surveys[3, ]    
# equivalent to head_surveys <- head(surveys)
head_surveys <- surveys[1:6, ] 
```
: is a special function that creates numeric vectors of integers in increasing or decreasing order, test 1:10 and 10:1 for instance.

You can also exclude certain indices of a data frame using the “-” sign:
 
```{r Indexing2}
surveys[, -1]          # The whole data frame, except the first column
surveys[-(7:34786), ] # Equivalent to head(surveys)

```

Data frames can be subset by calling indices (as shown previously), but also by calling their column names directly:

```{r indexing3}
surveys["species_id"]       # Result is a data.frame
surveys[, "species_id"]     # Result is a vector
surveys[["species_id"]]     # Result is a vector
surveys$species_id          # Result is a vector
```
In RStudio, you can use the autocompletion feature to get the full and correct names of the columns.

#Challenge Question 2
```{r Challenge 2}
#1
surveys_200 = surveys[200, ]
#2
rowCount = nrow(surveys)
surveysLastRow = surveys[rowCount, ]
#3
surveysMiddleRow = surveys[rowCount/2, ]
#4
surveys[-(7:rowCount), ]
```

## Factors

When we did str(surveys) we saw that several of the columns consist of integers. The columns genus, species, sex, plot_type, … however, are of a special class called factor. Factors are very useful and actually contribute to making R particularly well suited to working with data. So we are going to spend a little time introducing them.

Factors represent categorical data. They are stored as integers associated with labels and they can be ordered or unordered. While factors look (and often behave) like character vectors, they are actually treated as integer vectors by R. So you need to be very careful when treating them as strings.

Once created, factors can only contain a pre-defined set of values, known as levels. By default, R always sorts levels in alphabetical order. For instance, if you have a factor with 2 levels:

```{r factor 1}
sex <- factor(c("male", "female", "female", "male"))
```
R will assign 1 to the level "female" and 2 to the level "male" (because f comes before m, even though the first element in this vector is "male"). You can see this by using the function levels() and you can find the number of levels using nlevels():

```{r factor12}
levels(sex)
nlevels(sex)
```

Sometimes, the order of the factors does not matter, other times you might want to specify the order because it is meaningful (e.g., “low”, “medium”, “high”), it improves your visualization, or it is required by a particular type of analysis. Here, one way to reorder our levels in the sex vector would be:

```{r factor 2}
sex
sex <- factor(sex, levels = c("male", "female"))
sex
```
In R’s memory, these factors are represented by integers (1, 2, 3), but are more informative than integers because factors are self describing: "female", "male" is more descriptive than 1, 2. Which one is “male”? You wouldn’t be able to tell just from the integer data. Factors, on the other hand, have this information built in. It is particularly helpful when there are many levels (like the species names in our example datasset).

## Converting factors

If you need to convert a factor to a character vector, you use as.character(x).
```{r factor 3}
as.character(sex)

```

In some cases, you may have to convert factors where the levels appear as numbers (such as concentration levels or years) to a numeric vector. For instance, in one part of your analysis the years might need to be encoded as factors (e.g., comparing average weights across years) but in another part of your analysis they may need to be stored as numeric values (e.g., doing math operations on the years). This conversion from factor to numeric is a little trickier. The as.numeric() function returns the index values of the factor, not its levels, so it will result in an entirely new (and unwanted in this case) set of numbers. One method to avoid this is to convert factors to characters, and then to numbers.

Another method is to use the levels() function. Compare:

```{r factor 4}
year_fct <- factor(c(1990, 1983, 1977, 1998, 1990))
as.numeric(year_fct)               # Wrong! And there is no warning...
as.numeric(as.character(year_fct)) # Works...
as.numeric(levels(year_fct))[year_fct]    # The recommended way.
```

## Renaming Factors
When your data is stored as a factor, you can use the plot() function to get a quick glance at the number of observations represented by each factor level. Let’s look at the number of males and females captured over the course of the experiment:

```{r factor 5}
## bar plot of the number of females and males captured during the experiment:
plot(as.factor(surveys$sex))
```
In addition to males and females, there are about 1700 individuals for which the sex information hasn’t been recorded. Additionally, for these individuals, there is no label to indicate that the information is missing or undetermined. Let’s rename this label to something more meaningful. Before doing that, we’re going to pull out the data on sex and work with that data, so we’re not modifying the working copy of the data fra

```{r factor 6}
sex <- factor(surveys$sex)
head(sex)
levels(sex)
levels(sex)[1] <- "undetermined"
levels(sex)
head(sex)
```
# Challenge 3
```{r challenge 3}
# renaming F & M
levels(sex)[2:3] = c("female", "male")
#factoring levels for new DF
Challenge3DF =  factor(sex, levels = c("female", "male", "undetermined"))
plot(Challenge3DF)
```

### Using stringAsFactors = False

By default, when building or importing a data frame, the columns that contain characters (i.e. text) are coerced (= converted) into factors. Depending on what you want to do with the data, you may want to keep these columns as character. To do so, read.csv() and read.table() have an argument called stringsAsFactors which can be set to FALSE.

In most cases, it is preferable to set stringsAsFactors = FALSE when importing data and to convert as a factor only the columns that require this data type.
```{r reading 1}
## Compare the difference between our data read as `factor` vs `character`.
surveys <- read.csv("data/portal_data_joined.csv", stringsAsFactors = TRUE)
str(surveys)
surveys <- read.csv("data/portal_data_joined.csv", stringsAsFactors = FALSE)
str(surveys)
## Convert the column "plot_type" into a factor
surveys$plot_type <- factor(surveys$plot_type)
```

#Challenge 4
```{r challenge4-1}
#1
animal_data <- data.frame(
          animal = c("dog", "cat", "sea cucumber", "sea urchin"),
          feel = c("furry", 'furry', "squishy", "spiny"),
          weight = c(45, 8, 1.1, 0.8)
          )


#2
country_climate <- data.frame(
       country = c("Canada", "Panama", "South Africa", "Australia"),
       climate = c("cold", "hot", "temperate", "hot/temperate"),
       temperature = c(10, 30, 18, "15"),
       northern_hemisphere = c(TRUE, TRUE, FALSE, "FALSE"),
       has_kangaroo = c(FALSE, FALSE, FALSE, 1)
       )

str(country_climate)
```
for question 2, it would appear that all of the string columns in this data frame are factors, which is not what I would have expected for some of these fields (especially ones such as country and temperature). to fix these, I would stat by removing the quotes in the northern hemisphere  and temperature fields to make them numeric. Next I would then replace the final 1 in "has_kangaroo" with either true or false (in this case true since it is a 1). I would have made StringsAsFactors equal to flase in order to get character vectors over the factors that we see here.

_____________________________________________________________
The automatic conversion of data type is sometimes a blessing, sometimes an annoyance. Be aware that it exists, learn the rules, and double check that data you import in R are of the correct type within your data frame. If not, use it to your advantage to detect mistakes that might have been introduced during data entry (for instance, a letter in a column that should only contain numbers).


## Formatting Dates
One of the most common issues that new (and experienced!) R users have is converting date and time information into a variable that is appropriate and usable during analyses. As a reminder from earlier in this lesson, the best practice for dealing with date data is to ensure that each component of your date is stored as a separate variable. Using str(), We can confirm that our data frame has a separate column for day, month, and year, and that each contains integer values.
```{r dates 1}
library("lubridate")
library("tidyverse")
str(surveys)

```
We are going to use the ymd() function from the package lubridate (which belongs to the tidyverse; learn more here). lubridate gets installed as part as the tidyverse installation. When you load the tidyverse (library("tidyverse")), the core packages (the packages used in most data analyses) get loaded. lubridate however does not belong to the core tidyverse, so you have to load it explicitly with library(lubridate)

ymd() takes a vector representing year, month, and day, and converts it to a Date vector. Date is a class of data recognized by R as being a date and can be manipulated as such. The argument that the function requires is flexible, but, as a best practice, is a character vector formatted as “YYYY-MM-DD”.

Let’s create a date object and inspect the structure:
```{r dates 2}
my_date <- ymd("2015-01-01")
str(my_date)

# Now let’s paste the year, month, and day separately - we get the same result:
# sep indicates the character to use to separate each component
my_date <- ymd(paste("2015", "1", "1", sep = "-")) 
str(my_date)

#Now we apply this function to the surveys dataset. Create a character vector from the year, month, and day columns of surveys using paste():

paste(surveys$year, surveys$month, surveys$day, sep = "-")
```

This character vector can be used as the argument for ymd():
```{r dates 3}
ymd(paste(surveys$year, surveys$month, surveys$day, sep = "-"))
```
The resulting Date vector can be added to surveys as a new column called date:
```{r dates 4}
surveys$date <- ymd(paste(surveys$year, surveys$month, surveys$day, sep = "-"))

```


```{r dates 5}
str(surveys) # notice the new column, with 'date' as the class
```

Let’s make sure everything worked correctly. One way to inspect the new column is to use summary():

```{r dates 52}
summary(surveys$date)
```

Something went wrong: some dates have missing values. Let’s investigate where they are coming from.

We can use the functions we saw previously to deal with missing data to identify the rows in our data frame that are failing. If we combine them with what we learned about subsetting data frames earlier, we can extract the columns “year,”month“,”day" from the records that have NA in our new column date. We will also use head() so we don’t clutter the output:
```{r dates 6 }
missing_dates <- surveys[is.na(surveys$date), c("year", "month", "day")]

head(missing_dates)
```
Why did these dates fail to parse? If you had to use these data for your analyses, how would you deal with this situation?

