---
title: "Lab 5 : Data wrangling and graphing COVID-19 reporting data (Part 2)"
author: "Jarrod Daniels"
date: "9/21/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Joining tables

```{r library tidyverse}
library(tidyverse)
```

```{r load data 1}
Confirmed_State_3_13 <-   read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/03-13-2020.csv")) %>%
  rename(Country_Region = "Country/Region", Province_State = "Province/State") %>% 
  filter (Country_Region == "US") %>% 
  group_by(Province_State, Country_Region) %>% 
  summarise(Confirmed = sum(Confirmed)) 
```
```{r str confirmed 3_13}
str(Confirmed_State_3_13)
```

```{r confirmed 9_13}
Confirmed_State_9_13 <-   read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/09-13-2020.csv")) %>% 
  filter (Country_Region == "US") %>% 
  group_by(Province_State, Country_Region) %>% 
  summarise(Confirmed = sum(Confirmed)) 
```

```{r confirmed new}
str(Confirmed_State_9_13)
```
Notice from the above `str’ calls that there are 53 states in the table on 3/13/2020 and 58 states in the table on 9/13/2020

```{r}
setdiff(Confirmed_State_9_13$Province_State, Confirmed_State_3_13$Province_State)
```
There are 4 US territories a field you may have notice last week “Recovered”. We don’t need the “Recovered” row so let’s filter it from the data set.

```{r confrimed_state}
Confirmed_State_9_13 <- Confirmed_State_9_13 %>% 
  filter(Province_State != "Recovered") 
```

Since all of the states and territories in 9/13/2020 are found in 3/13/2020 a left_join or full_join can be used to merge the tables

```{r full join}
Confirmed_State_3_13_9_13_joined <- full_join(Confirmed_State_3_13,
      Confirmed_State_9_13, by = c("Province_State"))
head(Confirmed_State_3_13_9_13_joined)
```

This results in the values for Guam, Northern Mariana Islands, Puerto Rico and the Virgin Islands being NA for 3/13

```{r}
tail(Confirmed_State_3_13_9_13_joined, 5)
```
Or
```{r}
which(is.na(Confirmed_State_3_13_9_13_joined))
```

Let’s turn the NAs into 0. Also notice that we know have columns Confirmed.x Country_Region.x, Confirmed.y Country_Region.y. So let’s rename the Confirmed columns and delete the Country_Region columns

```{r}
Confirmed_State_3_13_9_13_joined <- full_join(Confirmed_State_3_13,
      Confirmed_State_9_13, by = c("Province_State")) %>% 
      rename(Confirmed_3_13_2020 = "Confirmed.x", Confirmed_9_13_2020 = "Confirmed.y") %>% 
      select(-Country_Region.x, -Country_Region.y) %>% 
      replace_na(list(Confirmed_3_13_2020 = 0))
head(Confirmed_State_3_13_9_13_joined)
```

```{r}
which(is.na(Confirmed_State_3_13_9_13_joined))
```

## Switching between wide and long table formats

gather and spead were used in the Data Carpentries tutorial to switch between wide and long table formats. They work fine, but are being replaced by the newer functions pivot_longer and pivot_wider. which we will use below. See the pivot vignette for other examples


```{r}
Confirmed_State_3_13_9_13_joined_long <- Confirmed_State_3_13_9_13_joined %>% 
              pivot_longer(-c(Province_State),
                            names_to = "Date", values_to = "Confirmed")
```

Now we can use this table in plot that shows the confirmed cases for both dates.

```{r, fig.width= 5, fig.height = 10}
# to adjust figure size {r, fig.width = 5, fig.height = 10}
ggplot(Confirmed_State_3_13_9_13_joined_long, aes(x = Confirmed,  y = Province_State))  + 
    geom_point(aes(color = Date)) 
```

## Working with time series data
# Data Wranling
Start by loading the most recent times series data for confirmed cases
```{r time series 1}
time_series_confirmed <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv")) %>%
  rename(Province_State = "Province/State", Country_Region = "Country/Region")
```

Note that in some cases it is best to have the date on your computer and then load the file into R.

```{r, eval=FALSE}
 download.file(url="https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv", 
               destfile = "data/time_series_covid19_confirmed_global.csv")
```

```{r}
time_series_confirmed <- read_csv("data/time_series_covid19_confirmed_global.csv")%>%
  rename(Province_State = "Province/State", Country_Region = "Country/Region")
```
Check the table properties to make sure the data imported as we expected

```{r}
head(time_series_confirmed)
```

As noted above this data is in wide format. To convert to long format

```{r}
time_series_confirmed_long <- time_series_confirmed %>% 
               pivot_longer(-c(Province_State, Country_Region, Lat, Long),
                            names_to = "Date", values_to = "Confirmed") 
```

Let’s look at the format of the data frame (tibble) now

```{r}
head(time_series_confirmed_long)
```

It would be convenient to have the confirmed cases and deaths in the same table. We can create another table with the deaths and then join the two tables. First let’s

```{r, eval=FALSE}
 download.file(url="https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv", 
               destfile = "data/time_series_covid19_deaths_global.csv")

```

```{r}
time_series_deaths <- read_csv("data/time_series_covid19_deaths_global.csv")%>%
  rename(Province_State = "Province/State", Country_Region = "Country/Region")
```


As noted above this data is in wide format. To convert to long format

```{r}
time_series_deaths_long <- time_series_deaths %>% 
               pivot_longer(-c(Province_State, Country_Region, Lat, Long),
                            names_to = "Date", values_to = "Deaths") 
head(time_series_deaths_long)

```

# Joining the time series tables

To join 2 tables we need a common column in which each row is a unique name. This is often called the Key. None of the columns in this data set meet that criteria, but we can create a column to serve as our key by mergering the names for the Province_State, Country_Region and Date columns using unite

```{r join 1}
time_series_confirmed_long <- time_series_confirmed_long %>% 
  unite(Key, Province_State, Country_Region, Date, sep = ".", remove = FALSE)
head(time_series_confirmed_long)
```

For the second table we can do the same and eliminate the columns that are redundant so that we just have the Key and Deaths columns

```{r}
time_series_deaths_long <- time_series_deaths_long %>% 
  unite(Key, Province_State, Country_Region, Date, sep = ".") %>% 
  select(Key, Deaths)
```

Now to join the tables. A full_join which means that the keys that are common to the tables will be joined and there will be counts for both Confirmed and Deaths. Where there are not matching values, returns NA for the one missing. select can be used to remove the key after the join since we don’t have further need for it.

```{r}

time_series_long_joined <- full_join(time_series_confirmed_long,
              time_series_deaths_long, by = c("Key")) %>% 
              select(-Key)
head(time_series_long_joined)

```

Check to make sure neither the Cofirmed or Death counts have NA as a value (both tables have the same number of rows, but we should check to make sure)

```{r}
which(is.na(time_series_long_joined$Confirmed))
```

```{r}
which(is.na(time_series_long_joined$Deaths))
```

```{r}
library(lubridate)
```

```{r}
time_series_long_joined$Date <- mdy(time_series_long_joined$Date)
```

Confirmed and Deaths are count data and there may be times when we want to plot both on the same graph. There are multiple ways to accomplish this and one is through creating a column using _pivot_long

```{r}
time_series_long_joined_counts <- time_series_long_joined %>% 
  pivot_longer(-c(Province_State, Country_Region, Lat, Long, Date),
               names_to = "Report_Type", values_to = "Counts")
head(time_series_long_joined_counts)
```

# Making Graphs from the time series data
To make a times series graph of the confirmed cases we need to summarize the Country date to count up the individual state data for the US:

```{r}
time_series_long_joined %>% 
  group_by(Country_Region,Date) %>% 
  summarise_at(c("Confirmed", "Deaths"), sum) %>% 
  filter (Country_Region == "US") %>% 
    ggplot(aes(x = Date,  y = Deaths)) + 
    geom_point() +
    geom_line() +
    ggtitle("US COVID-19 Deaths")
```

Let’s look at the US data in the context of a few other countries

```{r}
time_series_long_joined %>% 
  group_by(Country_Region,Date) %>% 
  summarise_at(c("Confirmed", "Deaths"), sum) %>% 
  filter (Country_Region %in% c("China","Japan", "Korea, South",
                                "Italy","Spain", "US")) %>% 
    ggplot(aes(x = Date,  y = Deaths)) + 
    geom_point() +
    geom_line() +
    ggtitle("COVID-19 Deaths") +
    facet_wrap(~Country_Region, ncol=2, scales="free_y")
```

Now several countries on the same graph
```{r}
time_series_long_joined %>% 
    group_by(Country_Region,Date) %>% 
    summarise_at(c("Confirmed", "Deaths"), sum) %>% 
    filter (Country_Region %in% c("China","France","Italy", 
                                "Korea, South", "US")) %>% 
    ggplot(aes(x = Date,  y = Deaths, color = Country_Region)) + 
    geom_point() +
    geom_line() +
    ggtitle("COVID-19 Deaths")
```

We can use the alternative data frame with column Report_Type to show Confirmed and Deaths. Because the counts for Deaths are low relative to Confirmed a log scale works best here for seeing the pattern

```{r}
time_series_long_joined_counts %>% 
  group_by(Country_Region, Report_Type, Date) %>% 
  summarise(Counts = sum(Counts)) %>% 
  filter (Country_Region == "US") %>% 
    ggplot(aes(x = Date,  y = log2(Counts), fill = Report_Type, color = Report_Type)) + 
    geom_point() +
    geom_line() +
    ggtitle("US COVID-19 Cases")
```

## Exercises 

```{r, fig.height= 10}
ggplot(Confirmed_State_3_13_9_13_joined_long, aes(y = Confirmed,  x = Province_State))  + 
    geom_bar(stat = "identity", aes(color = Date)) + coord_flip() + ylab("Confirmed Cases of Covid 19") + xlab("US State/Territory") + ggtitle("Confirmed Covid 19 Cases on 3/13/2020 vs 9/13/2020")
```
Just as a note, due to how small the numbers for 3/13 are in comparison 9/13, I have decided to keep the barplot stacked for the sake of visual clarity.


```{r}
time_series_long_joined %>% 
    group_by(Date) %>% 
    summarise_at(c("Confirmed", "Deaths"), sum) %>%
     ggplot(aes(x = Date,  y = Deaths)) + 
      geom_point() +
      geom_line() +
      ggtitle("Confirmed Covid 19 Deaths Around the Globe")
```
 
```{r}
t2 = time_series_long_joined %>% 
  mutate(Deaths_to_Confirmed_ratio = Deaths/Confirmed )


t2 %>% 
  filter (Country_Region == "US") %>% 
  ggplot(aes(x = Date,  y = Deaths_to_Confirmed_ratio)) + 
      geom_point() +
      geom_line() +
      ggtitle("Time Lapse of Covid-19 Death/Confirmed Ratio in the United States") + ylab("Death/Confirmed Cases")

```

```{r}
report_09_20_2020 <-   read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/09-12-2020.csv"))
```

```{r}
 Top10deaths = report_09_20_2020 %>% 
  group_by(Country_Region) %>% 
  summarise(Deaths = sum(Deaths)) %>% 
  arrange(desc(Deaths)) %>% 
  select(Country_Region) %>% 
  slice(0:10)

time_series_long_joined %>% 
    group_by(Country_Region,Date) %>% 
    summarise_at(c("Confirmed", "Deaths"), sum) %>% 
    filter (Country_Region %in% Top10deaths$Country_Region) %>% 
    ggplot(aes(x = Date,  y = Deaths, color = Country_Region)) + 
    geom_point() +
    geom_line() +
    ggtitle("Time Lapse of the 10 Countries with Greatest Number of Covid-19 Related Deaths")
```

```{r, fig.height= 10}
library(ggthemes)

time_series_long_joined %>% 
    group_by(Country_Region,Date) %>% 
    summarise_at(c("Confirmed", "Deaths"), sum) %>% 
    filter (Country_Region %in% Top10deaths$Country_Region) %>% 
    ggplot(aes(x = Date,  y = Deaths, color = Country_Region)) + 
    geom_point() +
    geom_line() +
    ggtitle("Time Lapse of the 10 Countries with Greatest Number of Covid-19 Related Deaths") +
    facet_wrap(~Country_Region, ncol=2, scales="free_y") +
    theme_economist() + scale_fill_economist()
    
  
```

```{r, message=FALSE, fig.height=15, fig.width= 13}
time_series_confirmed_US = read_csv("data/time_series_covid19_confirmed_US.csv")

time_series_confirmed_US = time_series_confirmed_US %>% 
               pivot_longer(-c(Province_State, Country_Region, UID, iso2, iso3, code3, FIPS, Admin2, Province_State, Country_Region, Lat, Long_, Combined_Key),
                            names_to = "Date", values_to = "Confirmed") 
time_series_confirmed_US$Date = mdy(time_series_confirmed_US$Date)

time_series_confirmed_US %>% 
  group_by(Province_State, Date) %>% 
  summarise_at(c("Confirmed"), sum) %>% 
  ggplot(aes(x = Date,  y = Confirmed)) + 
    geom_point() +
    geom_line() +
    ggtitle("Time Lapse of Confrimed COVID-19 Cases Across the US & Territories") +
    facet_wrap(~Province_State, ncol=5, scales="free_y") + ylab("Confirmed Cases")
```






