---
title: 'Lab 9A: Programming in R - Regular Expressions'
author: "Jarrod Daniels"
date: "10/26/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 14.3 Matchings patterns with regular expressions

Regexps are a very terse language that allow you to describe patterns in strings. They take a little while to get your head around, but once you understand them, you’ll find them extremely useful.

To learn regular expressions, we’ll use str_view() and str_view_all(). These functions take a character vector and a regular expression, and show you how they match. We’ll start with very simple regular expressions and then gradually get more and more complicated. Once you’ve mastered pattern matching, you’ll learn how to apply those ideas with various stringr functions.

```{r }
library(stringr)
library(tidyverse)
```

## 14.3.1 Basic Matches

```{r}
x <- c("apple", "banana", "pear")
str_view(x, "pl")
```

The next step up in complexity is ., which matches any character (except a newline):

```{r}
str_view(x, ".a.")
```

Note that this only shows the first match in a string.
But if “.” matches any character, how do you match the character “.”? You need to use an “escape” to tell the regular expression you want to match it exactly, not use its special behaviour. Like strings, regexps use the backslash, \, to escape special behaviour. So to match an ., you need the regexp \.. Unfortunately this creates a problem. We use strings to represent regular expressions, and \ is also used as an escape symbol in strings. So to create the regular expression \. we need the string "\\.".

```{r}
# To create the regular expression, we need \\
dot <- "\\."

# But the expression itself only contains one:
writeLines(dot)

# And this tells R to look for an explicit .
x <- c("abc", "a.c", "bef")
str_view(x, "a\\.c")
```

If \is used as an escape character in regular expressions, how do you match a literal \? Well you need to escape it, creating the regular expression \\. To create that regular expression, you need to use a string, which also needs to escape \. That means to match a literal \ you need to write "\\\\" — you need four backslashes to match one!

```{r}
x <- "a\\b"
writeLines(x)
```
```{r}
str_view(x, "\\\\")
```

In this book, I’ll write regular expression as . and strings that represent the regular expression as "\\.".

### 14.3.1.1 Exercises

1. The reasoning behind why this won't work is because you need 4 consecutive back slashes to match a "\" using regular expressions. you need "\\" to create the initial regular expression. What we are matching is a string which means we need another backslash. finally the fourth back slash is used for the actual matching characater.

## 14.3.2 Anchors

By default, regular expressions will match any part of a string. It’s often useful to anchor the regular expression so that it matches from the start or end of the string. You can use:

* ^ to match the start of the string.
* $ to match the end of the string.

```{r}
x <- c("apple", "banana", "pear")
str_view(x, "^a")
```

```{r}
str_view(x, "a$")
```

To remember which is which, try this mnemonic which I learned from Evan Misshula: if you begin with power (^), you end up with money ($).

To force a regular expression to only match a complete string, anchor it with both ^ and $:

```{r}
x <- c("apple pie", "apple", "apple cake")
str_view(x, "apple")
```

```{r}
str_view(x, "^apple$")
```

You can also match the boundary between words with \\\\b. I don’t often use this in R, but I will sometimes use it when I’m doing a search in RStudio when I want to find the name of a function that’s a component of other functions. For example, I’ll search for \\\\bsum\\\\b" to avoid matching summarise, summary, rowsum and so on.

### 14.3.2.1 Exercises

1. 

```{r}
str_view(c("$^$", "Jarrod$^$Daniels"), "^\\$\\^\\$$")
```

Since ^ and $ are both characters used in regex expressiong we need to escape the string while also anchoring it as well resulting in the expression above.

## 14.3.3 Character classes and alternatives

There are a number of special patterns that match more than one character. You’ve already seen ., which matches any character apart from a newline. There are four other useful tools:

* \d : matches any digit.
* \s : matches any whitespace (e.g. space, tab, newline).
* [abc] : matches a, b, or c.
* [^abc] : matches anything except a, b, or c.

Remember, to create a regular expression containing \\d or \\s, you’ll need to escape the \\ for the string, so you’ll type "\\\\d" or "\\\\s".

A character class containing a single character is a nice alternative to backslash escapes when you want to include a single metacharacter in a regex. Many people find this more readable.

```{r}
# Look for a literal character that normally has special meaning in a regex
x <- c("abc", "a.c", "a*c", "a c")
str_view(x, "a[.]c")
```

```{r}
str_view(x, ".[*]c")
```

```{r}
str_view(x, "a[ ]")
```

This works for most (but not all) regex metacharacters: $ . | ? * + ( ) [ {. Unfortunately, a few characters have special meaning even inside a character class and must be handled with backslash escapes: ] \ ^ and -.

You can use alternation to pick between one or more alternative patterns. For example, abc|d..f will match either ‘“abc”’, or “deaf”. Note that the precedence for | is low, so that abc|xyz matches abc or xyz not abcyz or abxyz. Like with mathematical expressions, if precedence ever gets confusing, use parentheses to make it clear what you want:

```{r}
x <- c("grey", "gray")
str_view(x, "gr(e|a)y")
```


### 14.3.3.1 Excercises

1.
a:
```{r}

#while the str view function is helpful, for this question is may be better to use the str_subset function

x = c("apple", "papple", "pear", "ear")
str_subset(x, "^[aeiou]")
```

b:

```{r}
str_subset(x, "^[^aeiou]")
```

c:
```{r}
x = c("Red", "reed", "kled", "kleed")
str_subset(x, "[^e]ed$")
```

d:

```{r}
x = c("learning", "rise", "guys", 'smile')
str_subset(x, "(ing|ise)$")
```

## 14.3.4 Repetition

The next step up in power involves controlling how many times a pattern matches:

* ?: 0 or 1
* +: 1 or more
* *: 0 or more

```{r}
x <- "1888 is the longest year in Roman numerals: MDCCCLXXXVIII"
str_view(x, "CC?")
```

```{r}
str_view(x, "CC+")
```

```{r}
str_view(x, 'C[LX]+')
```

Note that the precedence of these operators is high, so you can write: colou?r to match either American or British spellings. That means most uses will need parentheses, like bana(na)+.

You can also specify the number of matches precisely:

* {n}: exactly n
* {n,}: n or more
* {,m}: at most m
* {n,m}: between n and m

```{r}
str_view(x, "C{2}")
```

```{r}
str_view(x, "C{2,}")
```

By default these matches are “greedy”: they will match the longest string possible. You can make them “lazy”, matching the shortest string possible by putting a ? after them. This is an advanced feature of regular expressions, but it’s useful to know that it exists:

```{r}
str_view(x, 'C{2,3}?')
```

```{r}
str_view(x, 'C[LX]+?')
```

### 14.3.4.1 Exercise

In this case we coud say that...
? would translate to {0,1} in {m,n} form which corresponds to matching at most 1
+ would translate to {1,} which corresponds to matching 1 or more
* translates to {0,} which means matching 0 or more

## 14.3.5 Grouping and backreferences

Earlier, you learned about parentheses as a way to disambiguate complex expressions. Parentheses also create a numbered capturing group (number 1, 2 etc.). A capturing group stores the part of the string matched by the part of the regular expression inside the parentheses. You can refer to the same text as previously matched by a capturing group with backreferences, like \\1, \\2 etc. For example, the following regular expression finds all fruits that have a repeated pair of letters.

```{r}
str_view(fruit, "(..)\\1", match = TRUE)
```

### 14.3.5.1 Exercises

1
* three identical characters appearing in a row
* A set of characters proceeded by the same set of chars but in reverse order
* 2 characters that are repeated
* Char followed by another char (can be anything), the proceeded by the original char and another character (can be any) with the origianl character again.
* Three chars followed by at least 0 chars of any kind followed by the original characters in reverse order

## 14.4 Tools

Now that you’ve learned the basics of regular expressions, it’s time to learn how to apply them to real problems. In this section you’ll learn a wide array of stringr functions that let you:

* Determine which strings match a pattern.
* Find the positions of matches.
* Extract the content of matches.
* Replace matches with new values.
* Split a string based on a match.

## 14.4.1 Detect Matches 

To determine if a character vector matches a pattern, use str_detect(). It returns a logical vector the same length as the input:
```{r}
x <- c("apple", "banana", "pear")
str_detect(x, "e")
```

Remember that when you use a logical vector in a numeric context, FALSE becomes 0 and TRUE becomes 1. That makes sum() and mean() useful if you want to answer questions about matches across a larger vector:

```{r}
# How many common words start with t?
sum(str_detect(words, "^t"))

# What proportion of common words end with a vowel?
mean(str_detect(words, "[aeiou]$"))
```

When you have complex logical conditions (e.g. match a or b but not c unless d) it’s often easier to combine multiple str_detect() calls with logical operators, rather than trying to create a single regular expression. For example, here are two ways to find all words that don’t contain any vowels:

```{r}
# Find all words containing at least one vowel, and negate
no_vowels_1 <- !str_detect(words, "[aeiou]")

# Find all words consisting only of consonants (non-vowels)
no_vowels_2 <- str_detect(words, "^[^aeiou]+$")
identical(no_vowels_1, no_vowels_2)
```


The results are identical, but I think the first approach is significantly easier to understand. If your regular expression gets overly complicated, try breaking it up into smaller pieces, giving each piece a name, and then combining the pieces with logical operations.

A common use of str_detect() is to select the elements that match a pattern. You can do this with logical subsetting, or the convenient str_subset() wrapper:

```{r}
words[str_detect(words, "x$")]

str_subset(words, "x$")
```

Typically, however, your strings will be one column of a data frame, and you’ll want to use filter instead:
```{r}
df <- tibble(
  word = words, 
  i = seq_along(word)
)
df %>% 
  filter(str_detect(word, "x$"))
```

A variation on str_detect() is str_count(): rather than a simple yes or no, it tells you how many matches there are in a string:

```{r}
x <- c("apple", "banana", "pear")
str_count(x, "a")

# On average, how many vowels per word?
mean(str_count(words, "[aeiou]"))

```

It’s natural to use str_count() with mutate():

```{r}
df %>% 
  mutate(
    vowels = str_count(word, "[aeiou]"),
    consonants = str_count(word, "[^aeiou]")
  )
```

Note that matches never overlap. For example, in "abababa", how many times will the pattern "aba" match? Regular expressions say two, not three:

```{r}
str_count("abababa", "aba")
```

```{r}
str_view_all("abababa", "aba")
```

Note the use of str_view_all(). As you’ll shortly learn, many stringr functions come in pairs: one function works with a single match, and the other works with all matches. The second function will have the suffix _all.

## 14.4.1.1 Exercises 

1. Words that start or end with x
```{r}
# 1 regex
x = c("xapple", "apple", "applex")
x[str_detect(x, "(^x|x$)")]

# multiple str detects
start = str_detect(x, "^x")
end = str_detect(x, "x$")
x[start | end]
```

2. Words that start with a vowel and end with consonant
```{r}
# 1 regex
x = c("eat", "heat", "art", "heart")
x[str_detect(x, "^[aeiou].*[^aeiou]$")]

# multiple str detects
start = str_detect(x, "^[aeiou]")
end = str_detect(x, "[^aeiou]$")
x[start & end]
```

3. words that contain at least one of each vowel
```{r}
## this one is much more difficult, as a result we will refer to the key example from the workflow

#Single statment
pattern <-
  cross(rerun(5, c("a", "e", "i", "o", "u")),
    .filter = function(...) {
      x <- as.character(unlist(list(...)))
      length(x) != length(unique(x))
    }
  ) %>%
  map_chr(~str_c(unlist(.x), collapse = ".*")) %>%
  str_c(collapse = "|")

## What is happening here is using the rerun function, we have to check for each possible combination of vowels that could occur within a word (for instance eio would be different from oie and so forth)

# multiple string detects
words[str_detect(words, "a") &
  str_detect(words, "e") &
  str_detect(words, "i") &
  str_detect(words, "o") &
  str_detect(words, "u")]

#There do not appear to be any words that match this pattern
```

## 14.4.2 Extract matches

To extract the actual text of a match, use str_extract(). To show that off, we’re going to need a more complicated example. I’m going to use the Harvard sentences, which were designed to test VOIP systems, but are also useful for practicing regexps. These are provided in stringr::sentences:

```{r}
length(sentences)
```

```{r}
head(sentences)
```

Imagine we want to find all sentences that contain a colour. We first create a vector of colour names, and then turn it into a single regular expression:

```{r}
colours <- c("red", "orange", "yellow", "green", "blue", "purple")
colour_match <- str_c(colours, collapse = "|")
colour_match
```

Now we can select the sentences that contain a colour, and then extract the colour to figure out which one it is:

```{r}
has_colour <- str_subset(sentences, colour_match)
matches <- str_extract(has_colour, colour_match)
head(matches)
```

* The above example has an error

```{r}
has_colour
```

Note that str_extract() only extracts the first match. We can see that most easily by first selecting all the sentences that have more than 1 match:

```{r}
more <- sentences[str_count(sentences, colour_match) > 1]
str_view_all(more, colour_match)
```
```{r}
str_extract(more, colour_match)
```

This is a common pattern for stringr functions, because working with a single match allows you to use much simpler data structures. To get all matches, use str_extract_all(). It returns a list:

```{r}
str_extract_all(more, colour_match)
```

You’ll learn more about lists in lists and iteration.

If you use simplify = TRUE, str_extract_all() will return a matrix with short matches expanded to the same length as the longest:

```{r}
str_extract_all(more, colour_match, simplify = TRUE)
```

```{r}
x <- c("a", "a b", "a b c")
str_extract_all(x, "[a-z]", simplify = TRUE)
```

### 14.4.2.1. Exercises

1. 
To modify the expression we need to make sure that anything outside of the actual color itself is not being highlighted (exclude words that might contain the color within it)

this can be done using the double backslash followed by b which allows us to form word boundaries
```{r}
has_colour = str_c("\\b(", str_c(colours, collapse = "|"), ")\\b")


get = sentences[str_count(sentences, has_colour) > 1]
str_view_all(get, has_colour)
```

## 14.4.3 Grouped matches

Earlier in this chapter we talked about the use of parentheses for clarifying precedence and for backreferences when matching. You can also use parentheses to extract parts of a complex match. For example, imagine we want to extract nouns from the sentences. As a heuristic, we’ll look for any word that comes after “a” or “the”. Defining a “word” in a regular expression is a little tricky, so here I use a simple approximation: a sequence of at least one character that isn’t a space.

```{r}
noun <- "(a|the) ([^ ]+)"

has_noun <- sentences %>%
  str_subset(noun) %>%
  head(10)
has_noun %>% 
  str_extract(noun)
```

str_extract() gives us the complete match; str_match() gives each individual component. Instead of a character vector, it returns a matrix, with one column for the complete match followed by one column for each group:

```{r}
has_noun %>% 
  str_match(noun)
```

(Unsurprisingly, our heuristic for detecting nouns is poor, and also picks up adjectives like smooth and parked.)

If your data is in a tibble, it’s often easier to use tidyr::extract(). It works like str_match() but requires you to name the matches, which are then placed in new columns:

```{r}
tibble(sentence = sentences) %>% 
  tidyr::extract(
    sentence, c("article", "noun"), "(a|the) ([^ ]+)", 
    remove = FALSE
  )
```

Like str_extract(), if you want all matches for each string, you’ll need str_match_all().

### 14.4.3.1 Exercise
1. Find all words that come after a “number” like “one”, “two”, “three” etc. Pull out both the number and the word.

```{r}
## In this case we can just look for the specific number and see if it is proceeded by a string..

Zero_To_Ten = "\\b(zero|one|two|three|four|five|six|seven|eight|nine|ten) +(\\w+)"
sentences[str_detect(sentences, Zero_To_Ten)] %>%
  str_extract(Zero_To_Ten)

## to note: this is looking for numbers as strings, if we were looking for actual intergers or doubles this regex would not work

```

## 14.4.4 Replacing matches

str_replace() and str_replace_all() allow you to replace matches with new strings. The simplest use is to replace a pattern with a fixed string:
```{r}
x <- c("apple", "pear", "banana")
str_replace(x, "[aeiou]", "-")
```

```{r}
str_replace_all(x, "[aeiou]", "-")
```

With str_replace_all() you can perform multiple replacements by supplying a named vector:
```{r}
x <- c("1 house", "2 cars", "3 people")
str_replace_all(x, c("1" = "one", "2" = "two", "3" = "three"))
```

Instead of replacing with a fixed string you can use backreferences to insert components of the match. In the following code, I flip the order of the second and third words.

```{r}
sentences %>% 
  str_replace("([^ ]+) ([^ ]+) ([^ ]+)", "\\1 \\3 \\2") %>% 
  head(5)
```

### 14.4.4.1 

1. Replace all forward slashes in a string with backslashes.
```{r}
sentence2 = "Once fixed this // will become a backslash"

sentence2 %>%
  str_replace("//", "\\\\")
  
```

## 14.4.5 Splitting

Use str_split() to split a string up into pieces. For example, we could split sentences into words:

```{r}
sentences %>%
  head(5) %>% 
  str_split(" ")
```
Because each component might contain a different number of pieces, this returns a list. If you’re working with a length-1 vector, the easiest thing is to just extract the first element of the list:

```{r}
"a|b|c|d" %>% 
  str_split("\\|") %>% 
  .[[1]]
```

Otherwise, like the other stringr functions that return a list, you can use simplify = TRUE to return a matrix:

```{r}
sentences %>%
  head(5) %>% 
  str_split(" ", simplify = TRUE)
```

You can also request a maximum number of pieces:
```{r}
fields <- c("Name: Hadley", "Country: NZ", "Age: 35")
fields %>% str_split(": ", n = 2, simplify = TRUE)
```

Instead of splitting up strings by patterns, you can also split up by character, line, sentence and word boundary()s:

```{r}
x <- "This is a sentence.  This is another sentence."
str_view_all(x, boundary("word"))
```

```{r}
str_split(x, " ")[[1]]
```

```{r}
str_split(x, boundary("word"))[[1]]
```

### 14.4.5.1 Exercises

1. Split up a string like “apples, pears, and bananas” into individual components.
```{r}
word = "apples, pears, and bananas"
str_split(word, ", +(and +)?")[[1]]
```

## 14.7 stringi 

stringr is built on top of the stringi package. stringr is useful when you’re learning because it exposes a minimal set of functions, which have been carefully picked to handle the most common string manipulation functions. stringi, on the other hand, is designed to be comprehensive. It contains almost every function you might ever need: stringi has 250 functions to stringr’s 49.

If you find yourself struggling to do something in stringr, it’s worth taking a look at stringi. The packages work very similarly, so you should be able to translate your stringr knowledge in a natural way. The main difference is the prefix: str_ vs. stri_.
